Bayesian integration of networks without gold standards

### 2011-08-19 src/main/latex/paper/paper.tex
Incorporating Darren's changes in the paper

\section{Results:}
We present a novel algorithm for computing the probability of experimental observations to be correct that operates without gold standard reference data. We show that our algorithm even outperforms existing gold standard based methods. Finally, we apply the new algorithm on a large collection of genetic interaction and protein-protein interaction experiments.

@@

A significant portion of knowledge about molecular biological processes is distributed over a large number of online databases \citep{stein_creating_2002}. This knowledge has been obtained through experiments performed in laboratories all over the world. Overlaps often exists across the contents of these databases. The sub-discipline of integrative bioinformatics aims at collating this knowledge and making it accessible to both humans and computers. A large number of approaches to this problem exists. %\note{Mention some}.

A popular integration paradigm is the construction of functional networks \citep{von_mering_string:_2003,warde-farley_genemania_2010}\note{Also cite Katherine}. Functional networks represent several different types of relationships between biological entities in an abstract manner. Associations such as genetic interactions, protein-protein interactions, gene regulation and co-expression, are combined into abstract statements of functional relatedness, which are termed functional interactions.

@@

To provide a more generic solution to this problem, we present a fully Bayesian method that calculates, for each statement in a semantically-integrated dataset, the probability that it is true. We have evaluated the method's effectiveness in comparison to related methods. The validity of any results of the method's application to real data is hard to verify without knowing the absolute biological truth. Therefore, we have developed a tool that tests integration methods on simulated data with the same characteristics as real biological networks.

@@

The complete system of processes of a certain type within the cell (e.g. protein-protein interactions) can be modelled as a network $G=(V,E)$, where entities, such as proteins, are nodes (vertices) $V=\{v_1,\ldots,v_N\}$ and their associations are edges $E = \{e_1, \ldots,e_M\} \subseteq \binom{V}{2}$. If one considers each pair of entities to potentially have an edge, it is possible to model the process of prediction of such an edge by an experiment as follows. Let $X = \{X_1, \ldots, X_n \}$ be a vector of $n$ networks, which have been experimentally derived from $G$. Considering a single potential edge $e$, each experiment $X_i$ makes a statement about $e$'s existence. Let $D^e_i$ be a random variable that assumes realisation 1 when the $i$th experiment $X_i$ predicts that the edge exists and 0 when it predicts that the edge does not exist. Let $d^e_i$ be the measured realisation from $X_i$, then $(D^e_i=d^e_i)$ is the event that the measured realisation in experiment $i$ is taking place. Furthermore, let $D^e_{(n)}$ be the vector of all $n$ experimental measurement events $(D^e_i=d^e_i)$ for the edge. Finally, let $L^e$ be the event that the edge really does exist in $G$. We are interested in $\prob(L^e|D^e_{(n)})$, that is, the probability that the edge really exists given all our experimental measurements.

An important concept necessary to determine this probability is the Bayes factor \citep{goodman_toward_1999}. For each of the $n$ experiments a Bayes factor $\Lambda_i$ can be determined, which is defined as follows:

@@

However, instead of sampling from $\pi(\theta,G|X)$ directly, we employ Gibbs sampling \citep{hastings_monte_1970} and alternately sample from $\pi(\theta|G,X)$ and $\pi(G|\theta,X)$.

@@

The second step in each cycle is the sampling of a new error rate vector $\theta$ based on $G$. The error rates $\alpha_i$ and $\beta_i$ behave like the success rates for misreading each potential edge. Modelling each observation event over a potential edge as a Bernoulli experiment with such a success rate, the number of false positives and false negatives in an experimental graph $X_i$ would follow a binomial distribution. The success rates for a binomial distribution can be sampled using the Beta distribution. The Beta distribution takes two parameters, $a$ and $b$. The parameters $a$ and $b$ can be interpreted as one plus the number of successes and failures, respectively, in a binomial distributed dataset. In order to obtain values for $a$ and $b$, one can compare each $X_i$ to the currently assumed true graph $G$ and use it to count the number of supposed True Positives (tp), False Positives (fp), True Negatives (tn) and False Negatives (fn). Then, $\alpha$ and $\beta$ can be sampled as follows:
\begin{align}
  \alpha_i &\sim Be(\text{fp}_i+1, \text{tn}_i+1)\\
  \beta_i &\sim Be(\text{fn}_i+1, \text{tp}_i+1)
\end{align}

To initiate the algorithm as a whole, we need to generate initial error rate vector $\theta_{\text{init}}$. It is sufficient to sample the initial values for each $\alpha_i$ and $\beta_i$ from $Be(1,1)$, which is equivalent to the uniform distribution $\mathcal{U}_{[0,1]}$. Pseudocode for the complete algorithm is given in the supplementary material.

@@

In order to evaluate the method described above and to compare it against other probabilistic integration methods, a simulation and testing environment was created. The testing tool creates a random graph according to a specified topology. In the simulation, this graph assumes the role of a true biological graph. The tool then derives a set of graphs from the true graph with pre-determined error rates. In the simulation these graphs assume the role of experimental datasets.

@@

The simulated true graph is created as a scale-free graph using a preferential attachment algorithm, since most molecular-biological graphs have been shown to be belong to the scale-free class \citep{jeong_lethality_2001, eisenberg_preferential_2003}. A description of the algorithm can be found in the supplementary material.
\note{Mention how this does not match the algorithm's definition of the prior distribution of $G$, thus making it more challenging for the algorithm to perform.}%TODO

@@

After the simulated evidential graphs have been given to the integration method in question the resulting integrated network is evaluated against the original graph. There are a number of different quality measures that can be applied. One important aspect to measure the accuracy of the method to estimate the error rates in the individual experiments. We can define the quadratic loss regarding the error estimates as follows.
\begin{equation}
  L_\text{ER} = \frac{1}{2n}\sum_{i=1}^n (\alpha_i - \hat{\alpha_i})^2 + (\beta_i - \hat{\beta_i})^2
\end{equation}
where $\hat{\alpha_i}$ and $\hat{\beta_i}$ are the estimates of the false positive and false negative rates for experiment $i$ according to the integration method in question.

Also, to measure the accuracy of the final edge probabilities produced by the method, we can define further loss functions. Since we can expect a vast number of true negatives when working with sparse, scale-free graphs, it would be helpful to see the loss over interacting and non-interacting node pairs separately. These can be interpreted as analogue to the algorithms false negative rate and false positive rate.

\begin{align} 
	L^{(+)} &= \frac{1}{|E_G|}\sum_{e \in E_G} (1 - \prob(e \in E_G|X))^2\\
	L^{(-)} &= \frac{1}{|\binom{V}{2} \setminus E_G|}\sum_{e \in \binom{V}{2} \setminus E_G} \prob(e \in E_G|X)^2
\end{align}
Other performance metrics such as the area under ROC curves or weighted sums over the quadratic loss functions above are also possible to apply. However we find that the separate loss functions shown above give a more detailed picture of a method's behaviour.
\note{Maybe move to discussion. Discuss other possible scoring functions, e.g. weighted sums over FPs and FNs}

>>>>

\section{Results:}
We present a novel algorithm for computing the probability of experimental observations to be correct that operates without gold standard reference data. We show that our algorithm outperforms existing gold standard based methods. Finally, we apply the new algorithm on a large collection of genetic interaction and protein-protein interaction experiments.

@@

A significant proportion of knowledge about molecular biological processes is distributed over a large number of online databases \citep{stein_creating_2002}. This knowledge has been obtained through experiments performed in laboratories all over the world. Overlaps often exists across the contents of these databases. The sub-discipline of integrative bioinformatics aims at collating this knowledge and making it accessible to both humans and computers.

A popular integration paradigm is the construction of functional networks \citep{von_mering_string:_2003,warde-farley_genemania_2010}\note{Also cite Katherine}. Functional networks represent different types of relationships between biological entities in an abstract manner. Associations such as genetic interactions, protein-protein interactions, gene regulation and co-expression are combined into simple abstract statements of functional relatedness, which are termed functional interactions.

@@

To provide a more generic solution to this problem, we present a fully Bayesian method which calculates, for each statement in a semantically-integrated dataset, the probability that it is true. We have evaluated the method's effectiveness in comparison to related methods. The validity of any results of the method's application to real data is difficult to verify without knowing the absolute biological truth. Therefore, we have developed a tool that tests integration methods on simulated data with the same characteristics as real biological networks.

@@

The complete system of processes of a certain type within the cell (e.g. protein-protein interactions) can be modelled as a network $G=(V,E)$, where entities, such as proteins, are nodes (vertices) $V=\{v_1,\ldots,v_N\}$ and their associations are edges $E = \{e_1, \ldots,e_M\} \subseteq \binom{V}{2}$. If one considers each pair of entities to potentially have an edge, it is possible to model the process of prediction of such an edge by an experiment as follows. Let $X = \{X_1, \ldots, X_n \}$ be a collection of $n$ networks which have been experimentally derived from $G$. Considering a single potential edge $e$, each experiment $X_i$ makes a statement about $e$'s existence. Let $D^e_i$ be a random variable that assumes realisation 1 when the $i$th experiment $X_i$ predicts that the edge exists and 0 when it predicts that the edge does not exist. Let $d^e_i$ be the measured realisation from $X_i$, then $(D^e_i=d^e_i)$ is the event that the measured realisation in experiment $i$ is taking place. Furthermore, let $D^e_{(n)}$ be the vector of all $n$ experimental measurement events $(D^e_i=d^e_i)$ for the edge. Finally, let $L^e$ be the event that the edge really does exist in $G$. We are interested in $\prob(L^e|D^e_{(n)})$, that is, the probability that the edge really exists given all our experimental measurements.

An important concept necessary to determine this probability is the Bayes factor \citep{kass_bayes_1995}. For each of the $n$ experiments a Bayes factor $\Lambda_i$ can be determined, which is defined as follows:

@@

However, instead of sampling from $\pi(\theta,G|X)$ directly, we employ Gibbs sampling \citep{gelfand_sampling-based_1990} and alternately sample from $\pi(\theta|G,X)$ and $\pi(G|\theta,X)$.

@@

The second step in each cycle is the sampling of a new error rate vector $\theta$ based on $G$. The error rates $\alpha_i$ and $\beta_i$ behave like the success rates for misreading each potential edge. Modelling each observation event over a potential edge as a Bernoulli experiment with such a success rate, the number of false positives and false negatives in an experimental graph $X_i$ would follow a binomial distribution. The success rates for a binomial distribution can be sampled using the Beta distribution. The Beta distribution takes two parameters, $a$ and $b$. The parameters $a$ and $b$ can be interpreted as one plus the number of successes and failures, respectively, in a binomial distributed dataset. In order to obtain values for $a$ and $b$, one can compare each $X_i$ to the currently assumed true graph $G$ and use it to count the number of supposed True Positives (tp), False Positives (fp), True Negatives (tn) and False Negatives (fn). Then, the full conditionals for $\alpha$ and $\beta$ can be sampled as follows:
\begin{align}
  \alpha_i &\sim Be(\text{fp}_i+1, \text{tn}_i+1)\\
  \beta_i &\sim Be(\text{fn}_i+1, \text{tp}_i+1)
\end{align}

To initiate the algorithm as a whole, we need to generate initial error rate vector $\theta_{\text{init}}$. It is sufficient to sample the initial values for each $\alpha_i$ and $\beta_i$ from their prior distribution.
%$Be(1,1)$, which is equivalent to the uniform distribution $\mathcal{U}_{[0,1]}$. 
Pseudocode for the complete algorithm is given in the supplementary material.

@@

In order to evaluate the method described above and to compare it against other probabilistic integration methods, a simulation and testing environment was created. The testing tool creates a random graph according to a specified model. In the simulation, this graph assumes the role of a true biological graph. The tool then derives a set of graphs from the true graph with pre-determined error rates. In the simulation these graphs assume the role of experimental datasets.

@@

The simulated true graph is created as a scale-free graph using a preferential attachment algorithm, since most molecular-biological graphs have been shown to be approximately scale-free \citep{jeong_lethality_2001, eisenberg_preferential_2003}. A description of the algorithm can be found in the supplementary material.
\note{Mention how this does not match the algorithm's definition of the prior distribution of $G$, thus making it more challenging for the algorithm to perform.}%TODO

@@

After the simulated evidential graphs have been given to the integration method in question the resulting integrated network is evaluated against the original graph. There are a number of different quality measures that can be applied. One important aspect is to measure the accuracy of the method in estimating the error rates in the individual experiments. We can define the quadratic loss for the error estimates as follows:
\begin{equation}
  L_\text{ER} = \frac{1}{2n}\sum_{i=1}^n (\alpha_i - \hat{\alpha_i})^2 + (\beta_i - \hat{\beta_i})^2
\end{equation},
where $\hat{\alpha_i}$ and $\hat{\beta_i}$ are the estimates of the false positive and false negative rates for experiment $i$ according to the integration method in question.

Also, to measure the accuracy of the final edge probabilities produced by the method, we can define further loss functions. Since we can expect a vast number of true negatives when working with sparse, scale-free graphs, it would be helpful to see the loss over interacting and non-interacting node pairs separately. These can be interpreted as analogue to the algorithm's false negative rate and false positive rate:

\begin{align} 
	L^{(+)} &= \frac{1}{|E_G|}\sum_{e \in E_G} (1 - \prob(e \in E_G|X))^2\\
	L^{(-)} &= \frac{1}{|\binom{V}{2} \setminus E_G|}\sum_{e \in \binom{V}{2} \setminus E_G} \prob(e \in E_G|X)^2
\end{align}.
Other performance metrics such as the area under ROC curves or weighted sums over the quadratic loss functions above are also possible to apply. However we find that the separate loss functions shown above give a more detailed picture of a method's behaviour.
\note{Maybe move to discussion. Discuss other possible scoring functions, e.g. weighted sums over FPs and FNs}
